Four decades of water quality change in the upper San Francisco Estuary\footnote{\Sexpr{vers}\newline The views expressed in this paper are those of the authors and do not necessarily reflect the views or policies of the U.S. Environmental Protection Agency.}

### 2017-03-03 sftrends_manu.Rnw
DIN eval back to clam case study

Trend analysis is a broad discipline that has been applied to time series for the interpretation of environmentally-relevant changes.  Direct evaluation of an observed time series is often insufficient, given that a long-term change can be masked by variation at shorter time scales or the observed variation represents the combined effects of many variables\cite{Oneill89,Levin92}. Climate, local, and regional environmental effects may act individually or together over time to impose a change on time series, such that methods that account for variation at different scales have been used for trend analysis \cite{Bhangu97,Champely97,Chang08,Halliday12}.  As a practical approach for water quality evaluation, trend analysis of ecosystem repsonse indicators often focuses on tracking the change in concentrations or loads of nutrients over many years. Response indicators can vary naturally with changing flow conditions and may also reflect long-term effects of management or policy changes. For example, \ac{chla} concentration as a measure of phytoplankton response to nutrient inputs can follow seasonal patterns with cyclical variation in temperature and light changes throughout each year, whereas annual trends can follow long-term variation in nutrient inputs to the system \cite{Cloern96,Cloern10}. Similarly, nutrient trends that vary with hydrologic loading also vary as a function of utilization rates by primary producers or decomposition processes \cite{Sakamoto89,Schultz08,Harding16}. Time series analysis of ecosystem response indicators must simultaneously consider effects of processes at multiple scales and interactions between variables of interest to develop a more comprehensive description of system change.

Appropriate methods for the analysis of change depend largely on the question of interest and characteristics of the environmental dataset.  Trend analyses for aquatic systems have traditionally focused on comparisons between discrete periods of time to estimate direction and magnitude of a trend using non-parametric tests \cite{Hirsch91,Esterby96}. Development of these conventional approaches addressed limitations in historical monitoring datasets related to infrequent sampling and relatively few years of continuous data. Increased availability of multi-decadal datasets, particularly for high profile environments, has accelerated development of trend analysis methods that leverage the descriptive potential of long-term time series from continuous monitoring programs \cite{Bowes09,Halliday12}. These methods are often data-driven where the parameterization of a simple functional model can change smoothly over time given that relationships between water quality variables and potential drivers are dynamic.  The \ac{wrtds} approach was developed under this context and has been used to characterize decadal trends in running-water systems \cite{Hirsch10,Sprague11,Medalie12,Hirsch14,Pellerin14,Zhang16}. This method has the potential to provide a spatially and temporally robust description of trends by fitting a dynamic model with parameters that change relative to the domain of interest.  More recently, the \ac{wrtds} method was adapted for trend analysis in tidal waters, with a focus on \ac{chla} trends in Tampa Bay \cite{Beck15} and the Patuxent River Estuary \cite{Beck17}, and tidally-influenced time series of dissolved oxygen from continuous sonde measurements \cite{Beck15b}. These studies have demonstrated the potential of \ac{wrtds} for trend analysis in tidal waters. Application to additional datasets could provide further insight into environmental dynamics of aquatic systems.

The \ac{sfe} on the Pacific Coast of the United States is one of the most prominent and culturally significant estuaries in the western hemisphere \cite{Cloern15}.  Background nutrient concentrations in the Bay often exceed those associated with excessive primary production, although ecosystem responses symptomatic of eutrophication have historically been infrequent.  Recent changes in response to additional stressors (e.g., variation in freshwater inputs/withdrawals, invasive species, climate change) suggests that recent conditions have not followed past trajectories and more subtle spatial and temporal variation could provide clues that describe underlying properties of this system \cite{Cloern12b}.  The unique ecological and social context of the region provides a valuable opportunity to gain insight into ecosystem properties that define water quality dynamics at different scales. In particular, the Sacramento - San Joaquin River Delta (hereafter `Delta') is a mosaic of inflows upstream of the \ac{sfe} that receives and processes inputs from the larger watershed \ac{sfe} \cite{Jassby00,Jassby02,Jassby08}. A comprehensive monitoring dataset has been collected at several fixed locations in the upper estuary and Delta for the last four decades \cite{Jabusch16}.  Moreover, nutrient dynamics in the Delta are inherently linked to flow variation from inputs, withdrawal, impoundments, and downstream transport \cite{Novick15}, suggesting that an approach that explicitly considers flow effects is critical for trend analysis.  To date, the regional monitoring dataset for the northern \ac{sfe}, including the Delta, is an under-utitilized data source and a comprehensive analysis with \ac{wrtds} could facilitate an understanding of historical and recent changes in water quality.

The goal of this study was to provide a comprehensive description of nutrient trends in the northern \ac{sfe} and Delta region to inform understanding of ecosystem response dynamics and potential causes of water quality change. We applied the newly-adapted method of weighted regression for tidal waters to describe nitrogen trends in  different spatial and temporal contexts. The specific objectives were to \begin{inparaenum}[1\upshape)]  
\item quantify and interpret trends over four decades at ten stations in the Delta, including annual, seasonal, and spatial changes in nitrogen analytes and response to flow variation, and
\item provide detailed descriptions of two case studies in the context of conceptual relationships modelled with \ac{wrtds}.
\end{inparaenum}
The second objective evaluated two specific water quality stations as additional case studies o demonstrate complexities with nutrient response to flow, effects of nutrient-related source controls on ambient conditions, and effects of biological invasion by benthic filter feeders on primary production. Although quantitative descriptions of change can be ends in themselves, the results provide a means to more detailed understanding of ecosystem properties.  Products derived from \ac{wrtds} can be used to inform additional analyses, such as water quality response after removing annual, seasonal, or flow effects.

@@

The Delta region drains a 200 thousand km$^2$ watershed into the \ac{sfe}, which is the largest estuary on the Pacific coast of North America.  The watershed provides drinking water to over 25 million people, including irrigation for 18 thousand km$^2$ of agricultural land from the Central Valley watershed.  Water enters \ac{sfe} through the Sacramento and San Joaquin rivers that have a combined inflow of approximately 28 km$^3$ per year, with the Sacramento accounting for 84\% of inflow to the Delta.  The \ac{sfe} system is divided into several sub-bays including the Delta, with Suisun Bay immediately downstream of the Delta, San Pablo Bay to the north, South Bay, and the Central Bay that drains to the Pacific Ocean through the Golden Gate (\cref{fig:delt_map}).  Water dynamics in the \ac{sfe} and Delta are governed by inflows from the watershed, tidal exchange with the Pacific Ocean, and water withdrawals for municipal and agricultural use \cite{Jassby00}.  Seasonally, inflows from the watershed peak in the spring and early summer from snowmelt, whereas consumption, withdrawals, and export have steadily increased from 1960 to present, but vary depending on inter-annual climate effects \cite{Cloern12b}. Notable drought periods have occurred from 1976-1977, 1987-1992, and recently from 2013-2015 \cite{Cloern15}.

@@

Quantitative descriptions of nutrient dynamics in the Delta are challenging given multiple sources and the volume of water that is exchanged with natural and anthropogenic processes.  A comprehensive evaluation using mass-balance models to describe nutrient dynamics in the Delta demonstrated that nitrogen enters the system in different forms and is processed at different rates before export or removal \cite{Novick15}. For example, a majority of ammonium entering the system during the summer is nitrified or assimilated, whereas a considerable percentage of total nitrogen load to the Delta is exported.  Although, the focus of our analysis is not to quantify sources or sinks of nitrogen species, a quantitative evaluation of long-term trends will provide a more comprehensive historical interpretation to hypothesize the effects of future changes in the context of known dynamics.  Nutrients in the Delta also vary with seasonal and annual changes in the delivery of water inflows, including water exports directly from the system \cite{Jassby00,Jassby08}.  Our analysis also explicitly accounts for the effects of flow changes on nutrient response to better understand variation both within the Delta and potential mechanisms of downstream tranport.

@@

Multi-decadal time series of nutrients and flow records were used to develop a quantitative description of nitrogen trends in the northern \ac{sfe} region.  The \ac{iep} is a consortium of state and federal agencies that have maintained the \ac{emp} in the Delta region since 1975 \cite{IEP13}.  The \ac{emp} collects monthly water quality samples at 19 stations in the Delta, Suisun Bay, and northeastern San Pablo Bay.  Water samples were collected using a Van Dorn sample, a submersible pump, or a flow through system depending on site.  All samples were processed with standard QA/QC at the California Department of Water Resources Bryte Laboratory in Sacramento \cite{IEP13}.  Nutrient time series were obtained from the \ac{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}) at ten discrete sampling stations from 1976 through 2013 (\cref{fig:delt_map}).  Stations were grouped by location in the study area for comparison: peripheral stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; interior stations D19, D26, D28; and Suisun stations D4, D6, and D7.  These stations were chosen based on continuity of the water quality time series and geographic location for understanding trends.  Time series were complete for all stations except for an approximate ten year gap from 1996-2014 for D19.  Data were minimally processed, with the exception of averaging replicates that occurred on the same day.  The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two). Less than 3\% of all observations were left-censored, although variation was observed between analytes and location.  The ammonium time series had the most censored observations at sites \Sexpr{cvals$Site_Code[1]} (\Sexpr{cvals$cper[1]}\% of all observations), \Sexpr{cvals$Site_Code[2]} (\Sexpr{cvals$cper[2]}\%), \Sexpr{cvals$Site_Code[3]} (\Sexpr{cvals$cper[3]}\%), \Sexpr{cvals$Site_Code[4]} (\Sexpr{cvals$cper[4]}\%), and \Sexpr{cvals$Site_Code[5]} (\Sexpr{cvals$cper[5]}\%).

@@

\noindent where $N$ is one of three nitrogen analytes, time $t$ is a continuous variable as decimal time to capture the annual or seasonal trend, and $Q$ is the flow variable (either flow or salinity depending on station).  The seasonal trend is modelled as a sinusoidal component to capture periodicity between years.  The \ac{wrtds} model is a moving window regression that fits unique parameters at each observation point in the time series.  A unique set of weights is used for each regression to control the importance of observations used to fit the model relative to the observation at the center of the window. The weights are based on a scaled Euclidean distance to estimate the differences of all points from the center in relation to annual time, season, and flow. The complete model for the time series contains a parameter set for every time step that considers the unique context of the data.  As such, predictions from \ac{wrtds} are more precise than those from more conventional models that fit a single parameter set to the entire time series \cite{Moyer12,Beck15}.  The \ac{wrtds} model applied to the Delta time series was based on a tidal adaptation of the original method\cite{Beck15}.  The \ac{wrtds} models were fit to describe the conditional mean response using a weighted Tobit model for left-censored data \cite{Tobin58}.  Previous adaptations of \ac{wrtds} to tidal waters have used quantile regression to describe trends in the conditional quantiles, such as changes in the frequency of occurrence of extreme events.  The application to the Delta data focused only on the conditional mean models to establish a baseline response which has not been previously quantified.  All analyses used the WRTDStidal package written by the authors for the R statistical programming language \cite{Beck16b,RDCT16}.

@@

The observed time series for the ten Delta stations had substantial variation in scale among the nitrogen analytes and differences in apparent seasonal trends (\cref{fig:obsdat}).  In general, long-term (inter-annual) trends were not easily observed from the raw data. \ac{din} for most stations was dominated by nitrite/nitrate, whereas ammonium was a smaller percentage of the total.  However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D26) had higher concentrations of ammonium during winter months when phytoplankton assimilation is lower\cite{Novick15}.  By location, observed concentrations of \ac{din} for the entire time series were higher on average for the peripheral stations (C3, C10, MD10, P8; maximum likelihood estimation of mean $\pm$ standard error: \Sexpr{scinot(obslo$Delta['mean'])}$\pm$\Sexpr{scinot(obslo$Delta['se'])} \mgl) and similar for the interior (D19, D26, D28, \Sexpr{scinot(obslo$Middle['mean'])}$\pm$\Sexpr{scinot(obslo$Middle['se'], 3)}) and Suisun Bay stations (D4, D6, D7, \Sexpr{scinot(obslo$Suisun['mean'])}$\pm$\Sexpr{scinot(obslo$Suisun['se'], 3)}).  Average concentrations were highest at \Sexpr{obssi$din[1, 'Site_Code']} (\Sexpr{scinot(obssi$din[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$din[1, 'se'])} \mgl) and lowest at \Sexpr{obssi$din[2, 'Site_Code']} (\Sexpr{scinot(obssi$din[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$din[2, 'se'], 3)}) for \ac{din}, highest at \Sexpr{obssi$nh[1, 'Site_Code']} (\Sexpr{scinot(obssi$nh[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$nh[1, 'se'])}) and lowest at \Sexpr{obssi$nh[2, 'Site_Code']} (\Sexpr{scinot(obssi$nh[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$nh[2, 'se'], 3, 3)}) for ammonium, and highest at \Sexpr{obssi$no23[1, 'Site_Code']} (\Sexpr{scinot(obssi$no23[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$no23[1, 'se'])}) and lowest at \Sexpr{obssi$no23[2, 'Site_Code']} (\Sexpr{scinot(obssi$no23[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$no23[2, 'se'], 3, 3)}) for nitrite/nitrate. Mean observed concentrations were also higher later in the time series for all analytes.  For example, average \ac{din} across all stations was \Sexpr{scinot(obsyr$din[1, 'mean'])}$\pm$\Sexpr{scinot(obsyr$din[1, 'se'])} \mgl for \Sexpr{obsyr$din[1, 'yr']}, compared to \Sexpr{scinot(obsyr$din[2, 'mean'])}$\pm$\Sexpr{scinot(obsyr$din[2, 'se'])} for \Sexpr{obsyr$din[2, 'yr']}. Seasonal changes across all years showed that nitrogen concentrations were generally lower in the summer and higher in the winter, although observed patterns were inconsistent between sites.  For example, site MD10 had distinct seasonal spikes for elevated \ac{din} in the winter, whereas other stations had less prominent seasonal maxima (e.g., C3, D7, \cref{fig:obsdat}).

@@

Significant efforts have been made in recent years to reduce nitrogen loading from regional \acp{wwtp} given the disproportionate contribution of nutrients relative to other sources (e.g., watershed agricultural load, sediment flux, etc.)\cite{Cornwell14,Novick15}.  Several \acp{wwtp} in the Delta have recently been or are planned to be upgraded to include tertiary filtration and nitrification to convert biologically available ammonium to nitrate. The City of Stockton \ac{wwtp} was upgraded in 2006 and is immediately upstream of station P8 (\cref{fig:delt_map})\cite{Jabusch16}, which provides a valuable opportunity to assess how nutrient or nutrient-related source controls and water management actions have changed ambient concentrations downstream. A modal response of nutrient concentrations at P8 centered around 2006 is expected as a result of upstream \ac{wwtp} upgrades, and water quality should exhibit \begin{inparaenum}[1\upshape)]
\item a shift in nutrient contributions from the \ac{wwtp} before/after upgrade, and
\item a flow-normalized annual trend at P8 to show a change concurrent with \ac{wwtp} upgrades.
\end{inparaenum}

@@

Effluent concentrations measured from 2003 to 2009 from the Stockton \ac{wwtp} showed a gradual reduction in ammonium concentration relative to the total (\cref{fig:stock}).  Ammonium and nitrate concentrations were comparable prior to 2006, whereas nitrate was a majority of total nitrogen after the upgrade with much smaller percentages from ammonium and nitrite. As expected, flow-normalized nitrogen trends at P8 shifted in response to upstream \ac{wwtp} upgrades (\cref{fig:p8trnds}a), with ammonium showing an increase from 1976 followed by a large reduction in the 2000s.  Interestingly, nitrite/nitrate concentrations also showed a similar but less dramatic decrease despite an increase in the \ac{wwtp} effluent concentrations following the upgrade.  Percent changes from seasonal Kendall tests on flow-normalized results showed that both nitrogen species increased prior to \ac{wwtp} upgrades (\Sexpr{scinot(no_bf[1, 'perchg'], 1, 1)}\% per year for nitrite/nitrate, \Sexpr{scinot(nh_bf[1, 'perchg'], 1, 1)}\% for ammonium), followed by decreases after upgrades (\Sexpr{scinot(no_af[1, 'perchg'], 1, 1)}\% for nitrite/nitrate, \Sexpr{scinot(nh_af[1, 'perchg'], 1, 1)}\% for ammonium, \cref{tab:p8chg}).  Seasonally, increases prior to upgrades were highest in the summer for nitrite/nitrate (\Sexpr{scinot(no_bf[3, 'perchg'], 1, 1)}\%) and in the fall for ammonium (\Sexpr{scinot(nh_bf[4, 'perchg'], 1, 1)}\%). Similarly, seasonal reductions post-upgrade were largest in the summer for nitrite/nitrate (\Sexpr{scinot(no_af[3, 'perchg'], 1, 1)}\%) and largest for ammonium in the winter (\Sexpr{scinot(nh_af[5, 'perchg'], 1, 1)}\%).

@@

% \section{References} % omit on submit
\begin{singlespace}
\bibliography{refs}
\end{singlespace}

@@

<<clmchl, fig.height = 9.3, fig.width = 8, out.width = '0.9\\textwidth', cache = T, fig.cap = 'Trends in clam abundance and \\ac{chla} concentration from 1976 to 2013 at station D7 in Suisun Bay.  Invasion by \\textit{Potamocorbula amurensis} clams in the late 1980s and displacement of \\textit{Corbicula fluminea} was shown by changes in clam density (a, annual means), with biomass linked to salinity (b).  A decrease in \\ac{chla} concentration was also observed by changes in annual (c) and seasonal trends (d) based on \\ac{wrtds} results.  Reductions in \\ac{chla} concentration were coincident with an increase in SiO$_2$ concentration (e), with the greatest increase in August (f). A significant ($p < 0.001$) relationship between clam biomass and \\ac{chla} concentration is shown in subfigure (g).  Flow relationships with \\ac{chla} concentration shown by \\ac{wrtds} have also changed over time (h, observations from June).'>>=

>>>>

Trend analysis is a broad discipline that has been applied to time series for the interpretation of environmentally-relevant changes.  Direct evaluation of an observed time series is often insufficient, given that a long-term change can be masked by variation at shorter time scales or the observed variation represents the combined effects of many variables\cite{Oneill89,Levin92}. Climate, local, and regional environmental effects may act individually or together over time to impose a change on time series, such that methods that account for variation at different scales have been used for trend analysis \cite{Bhangu97,Champely97,Chang08,Halliday12}.  As a practical approach for water quality evaluation, trend analysis of ecosystem response indicators often focuses on tracking the change in concentrations or loads of nutrients over many years. Response indicators can vary naturally with changing flow conditions and may also reflect long-term effects of management or policy changes. For example, \ac{chla} concentration as a measure of phytoplankton response to nutrient inputs can follow seasonal patterns with cyclical variation in temperature and light changes throughout each year, whereas annual trends can follow long-term variation in nutrient inputs to the system \cite{Cloern96,Cloern10}. Similarly, nutrient trends that vary with hydrologic loading also vary as a function of utilization rates by primary producers or decomposition processes \cite{Sakamoto89,Schultz08,Harding16}. Time series analysis of ecosystem response indicators must simultaneously consider effects of processes at multiple scales and interactions between variables of interest to develop a more comprehensive description of system change.

Appropriate methods for the analysis of change depend largely on the question of interest and characteristics of the environmental dataset.  Trend analyses for aquatic systems have traditionally focused on comparisons between discrete periods of time to estimate direction and magnitude of a trend using non-parametric tests \cite{Hirsch91,Esterby96}. Development of these conventional approaches addressed limitations in historical monitoring datasets related to infrequent sampling and relatively few years of continuous data. Increased availability of multi-decadal datasets, particularly for high profile environments, has accelerated development of trend analysis methods that leverage the descriptive potential of long-term time series from continuous monitoring programs \cite{Bowes09,Halliday12}. These methods are often data-driven where the parameterization of a simple functional model can change smoothly over time given that relationships between water quality variables and potential drivers are dynamic.  The \ac{wrtds} approach was developed in this context and has been used to characterize decadal trends in running-water systems \cite{Hirsch10,Sprague11,Medalie12,Hirsch14,Pellerin14,Zhang16}. This method has the potential to provide a spatially and temporally robust description of trends by fitting a dynamic model with parameters that change relative to the domain of interest.  More recently, the \ac{wrtds} method was adapted for trend analysis in tidal waters, with a focus on \ac{chla} trends in Tampa Bay \cite{Beck15} and the Patuxent River Estuary \cite{Beck17}, and tidally-influenced time series of dissolved oxygen from continuous sonde measurements \cite{Beck15b}. These studies have demonstrated the potential of \ac{wrtds} for trend analysis in tidal waters. Application to additional datasets could provide further insight into environmental dynamics of aquatic systems.

The Sacramento - San Joaquin River Delta (hereafter `Delta') is a mosaic of inflows upstream of the \ac{sfe} that receives and processes inputs from the larger watershed \cite{Jassby00,Jassby02,Jassby08}. Sediment export downstream of the Delta and \ac{wwtp} inputs are primary sources of nutrients for the larger Bay.  Background nutrient concentrations in \ac{sfe} often exceed those associated with excessive primary production, although ecosystem responses symptomatic of eutrophication have historically been infrequent.  Changes in response to additional stressors (e.g., variation in freshwater inputs/withdrawals, invasive species, climate change) suggests that recent conditions have not followed past trajectories and more subtle spatial and temporal variation could provide clues that describe underlying properties of this system \cite{Cloern12b}.  A comprehensive monitoring dataset has been collected at several fixed locations in the upper estuary and Delta for the last four decades \cite{Jabusch16}.  Moreover, nutrient dynamics in the Delta are inherently linked to flow variation from inputs, withdrawal, impoundments, and downstream transport \cite{Novick15}, suggesting that an approach that explicitly considers flow effects is critical for trend analysis.  To date, the regional monitoring dataset for the northern \ac{sfe}, including the Delta, is an under-utitilized data source and a comprehensive analysis with \ac{wrtds} could facilitate an understanding of historical and recent changes in water quality.

The goal of this study was to provide a comprehensive description of nutrient trends in the northern \ac{sfe} and Delta region to inform understanding of ecosystem response dynamics and potential causes of water quality change. We applied the newly-adapted method of weighted regression for tidal waters to describe nitrogen trends in  different spatial and temporal contexts. The specific objectives were to \begin{inparaenum}[1\upshape)]  
\item quantify and interpret trends over four decades at ten stations in the Delta, including annual, seasonal, and spatial changes in nitrogen analytes and response to flow variation, and
\item provide detailed descriptions of two case studies in the context of conceptual relationships modeled with \ac{wrtds}.
\end{inparaenum}
The second objective evaluated two specific water quality stations as additional case studies to demonstrate complexities with nutrient response to flow, effects of nutrient-related source controls on ambient conditions, and effects of biological invasion by benthic filter feeders on primary production. Although quantitative descriptions of change can be ends in themselves, the results provide means to a more detailed understanding of ecosystem properties.  Products derived from \ac{wrtds} can be used to inform additional analyses, such as water quality response after removing annual, seasonal, or flow effects.

@@

The Delta region drains a 200 thousand km$^2$ watershed into the \ac{sfe}, which is the largest estuary on the Pacific coast of North America.  The watershed provides drinking water to over 25 million people and includes irrigation for 18 thousand km$^2$ of agricultural land.  Water enters the \ac{sfe} through the Sacramento and San Joaquin rivers that have a combined inflow of approximately 28 km$^3$ per year, with the Sacramento accounting for 84\% of inflow to the Delta.  The \ac{sfe} system includes the Delta and subembayments of San Francisco Bay, with Suisun Bay immediately downstream of the Delta, San Pablo Bay to the north, South Bay, and the Central Bay that drains to the Pacific Ocean through the Golden Gate (\cref{fig:delt_map}).  Water dynamics in the \ac{sfe} and Delta are governed by inflows from the watershed, tidal exchange with the Pacific Ocean, and water withdrawals for municipal and agricultural use \cite{Jassby00}.  Seasonally, inflows from the watershed peak in the spring and early summer from snowmelt, whereas consumption, withdrawals, and export have steadily increased from 1960 to present, but vary depending on inter-annual climate effects \cite{Cloern12b}. Notable drought periods have occurred from 1976-1977, 1987-1992, and recently from 2013-2015 \cite{Cloern15}.

@@

Quantitative descriptions of nutrient dynamics in the Delta are challenging given multiple sources and the volume of water that is exchanged with natural and anthropogenic processes.  A comprehensive evaluation using mass-balance models to describe nutrient dynamics in the Delta demonstrated that nitrogen enters the system in different forms and is processed at different rates before export or removal \cite{Novick15}. For example, a majority of ammonium entering the system during the summer is nitrified or assimilated, whereas a considerable percentage of total nitrogen load to the Delta is exported.  Although, the focus of our analysis is not to quantify sources or sinks of nitrogen species, a quantitative evaluation of long-term trends will provide a more comprehensive historical interpretation to hypothesize the effects of future changes in the context of known dynamics.  Nutrients in the Delta also vary with seasonal and annual changes in the delivery of water inflows and water exports directly from the system \cite{Jassby00,Jassby08}.  Our analysis explicitly accounts for the effects of flow changes on nutrient response to better understand variation both within the Delta and potential mechanisms of downstream tranport.

@@

Multi-decadal time series of nutrients and flow records were used to develop a quantitative description of nitrogen trends in the northern \ac{sfe} region.  The \ac{iep} is a consortium of state and federal agencies that have maintained the \ac{emp} in the Delta region since 1975 \cite{IEP13}.  The \ac{emp} collects monthly water quality samples at 19 stations in the Delta, Suisun Bay, and northeastern San Pablo Bay.  Water samples were collected using a Van Dorn sample, a submersible pump, or a flow through system depending on site.  All samples were processed with standard QA/QC at the California Department of Water Resources Bryte Laboratory in Sacramento \cite{IEP13}.  Nutrient time series from 1976 to 2013 were obtained from the \ac{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}) for ten discrete sampling stations (\cref{fig:delt_map}).  Stations were grouped by location in the study area for comparison: peripheral Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; interior Delta stations D19, D26, D28; and Suisun stations D4, D6, and D7.  These stations were chosen based on continuity of the water quality time series and significance of their geographic location for understanding regional trends.  Time series were complete for all stations except for an approximate ten year gap from 1996-2014 for D19.  Data were minimally processed, with the exception of averaging replicates that occurred on the same day.  The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two). Less than 3\% of all observations were left-censored, although variation was observed between analytes and location.  The ammonium time series had the most censored observations at sites \Sexpr{cvals$Site_Code[1]} (\Sexpr{cvals$cper[1]}\% of all observations), \Sexpr{cvals$Site_Code[2]} (\Sexpr{cvals$cper[2]}\%), \Sexpr{cvals$Site_Code[3]} (\Sexpr{cvals$cper[3]}\%), \Sexpr{cvals$Site_Code[4]} (\Sexpr{cvals$cper[4]}\%), and \Sexpr{cvals$Site_Code[5]} (\Sexpr{cvals$cper[5]}\%).

@@

\noindent where $N$ is one of three nitrogen analytes, time $t$ is a continuous variable as decimal time to capture the annual or seasonal trend, and $Q$ is the flow variable (either flow or salinity depending on station).  The seasonal trend is modeled as a sinusoidal component to capture periodicity between years.  The \ac{wrtds} model is a moving window regression that fits unique parameters at each observation point in the time series.  A unique set of weights is used for each regression to control the importance of observations used to fit the model relative to the observation at the center of the window. The weights are based on a scaled Euclidean distance to estimate the differences of all points from the center in relation to annual time, season, and flow. The complete model for the time series contains a parameter set for every time step that considers the unique context of the data.  As such, predictions from \ac{wrtds} are more precise than those from more conventional models that fit a single parameter set to the entire time series \cite{Moyer12,Beck15}.  The \ac{wrtds} models applied herein were based on a tidal adaptation of the original method\cite{Beck15} and were fit to describe the conditional mean response using a weighted Tobit model for left-censored data \cite{Tobin58}. All analyses used the WRTDStidal package written by the authors for the R statistical programming language \cite{Beck16b,RDCT16}.

@@

The observed time series for the ten Delta - Suisun Bay stations had substantial variation in scale among the nitrogen analytes and differences in apparent seasonal trends (\cref{fig:obsdat}).  In general, long-term (inter-annual) trends were not easily observed from the raw data. \ac{din} for most stations was dominated by nitrite/nitrate, whereas ammonium was a smaller percentage of the total.  However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D26) had higher concentrations of ammonium during winter months when phytoplankton assimilation is lower\cite{Novick15}.  By location, observed concentrations of \ac{din} for the entire time series were higher on average for the peripheral stations (C3, C10, MD10, P8; maximum likelihood estimation of mean $\pm$ standard error: \Sexpr{scinot(obslo$Delta['mean'])}$\pm$\Sexpr{scinot(obslo$Delta['se'])} \mgl) and similar for the interior (D19, D26, D28, \Sexpr{scinot(obslo$Middle['mean'])}$\pm$\Sexpr{scinot(obslo$Middle['se'], 3)}) and Suisun Bay stations (D4, D6, D7, \Sexpr{scinot(obslo$Suisun['mean'])}$\pm$\Sexpr{scinot(obslo$Suisun['se'], 3)}).  Average concentrations were highest at \Sexpr{obssi$din[1, 'Site_Code']} (\Sexpr{scinot(obssi$din[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$din[1, 'se'])} \mgl) and lowest at \Sexpr{obssi$din[2, 'Site_Code']} (\Sexpr{scinot(obssi$din[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$din[2, 'se'], 3)}) for \ac{din}, highest at \Sexpr{obssi$nh[1, 'Site_Code']} (\Sexpr{scinot(obssi$nh[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$nh[1, 'se'])}) and lowest at \Sexpr{obssi$nh[2, 'Site_Code']} (\Sexpr{scinot(obssi$nh[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$nh[2, 'se'], 3, 3)}) for ammonium, and highest at \Sexpr{obssi$no23[1, 'Site_Code']} (\Sexpr{scinot(obssi$no23[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$no23[1, 'se'])}) and lowest at \Sexpr{obssi$no23[2, 'Site_Code']} (\Sexpr{scinot(obssi$no23[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$no23[2, 'se'], 3, 3)}) for nitrite/nitrate. Mean observed concentrations were also higher later in the time series for all analytes.  For example, average \ac{din} across all stations was \Sexpr{scinot(obsyr$din[1, 'mean'])}$\pm$\Sexpr{scinot(obsyr$din[1, 'se'])} \mgl for \Sexpr{obsyr$din[1, 'yr']}, compared to \Sexpr{scinot(obsyr$din[2, 'mean'])}$\pm$\Sexpr{scinot(obsyr$din[2, 'se'])} for \Sexpr{obsyr$din[2, 'yr']}. Seasonal changes across all years showed that nitrogen concentrations were generally lower in the summer and higher in the winter, although observed patterns were inconsistent between sites.  For example, site MD10 had distinct seasonal spikes for elevated \ac{din} in the winter, whereas other stations had less prominent seasonal maxima (e.g., C3, D7, \cref{fig:obsdat}).

@@

Significant efforts have been made in recent years to reduce nitrogen loading from regional \acp{wwtp} given the disproportionate contribution of nutrients relative to other sources (e.g., watershed agricultural load, sediment flux, etc.)\cite{Cornwell14,Novick15}.  Several \acp{wwtp} in the Delta have recently been or are planned to be upgraded to include tertiary filtration and nitrification to convert biologically available ammonium to nitrate. The City of Stockton \ac{wwtp} was upgraded in 2006 and is immediately upstream of station P8 (\cref{fig:delt_map})\cite{Jabusch16}, which provides a valuable opportunity to assess how nutrient or nutrient-related source controls and water management actions have changed ambient concentrations downstream. A modal response of nutrient concentrations at P8 centered around 2006 is expected as a result of upstream \ac{wwtp} upgrades, and water quality should exhibit \begin{inparaenum}[1\upshape)]
\item a shift in the ratio of the components of \ac{din} from the \ac{wwtp} before/after upgrade, and
\item a flow-normalized annual trend at P8 to show a change concurrent with \ac{wwtp} upgrades.
\end{inparaenum}

@@

Effluent concentrations measured from 2003 to 2009 from the Stockton \ac{wwtp} showed a gradual reduction in ammonium concentration relative to total \ac{din} (\cref{fig:stock}).  Ammonium and nitrate concentrations were comparable prior to 2006, whereas nitrate was a majority of total nitrogen after the upgrade, with much smaller percentages from ammonium and nitrite. As expected, flow-normalized nitrogen trends at P8 shifted in response to upstream \ac{wwtp} upgrades (\cref{fig:p8trnds}a), with ammonium showing an increase from 1976 followed by a large reduction in the 2000s.  Interestingly, nitrite/nitrate concentrations also showed a similar but less dramatic decrease despite an increase in the \ac{wwtp} effluent concentrations following the upgrade.  Percent changes from seasonal Kendall tests on flow-normalized results showed that both nitrogen species increased prior to \ac{wwtp} upgrades (\Sexpr{scinot(no_bf[1, 'perchg'], 1, 1)}\% per year for nitrite/nitrate, \Sexpr{scinot(nh_bf[1, 'perchg'], 1, 1)}\% for ammonium), followed by decreases after upgrades (\Sexpr{scinot(no_af[1, 'perchg'], 1, 1)}\% for nitrite/nitrate, \Sexpr{scinot(nh_af[1, 'perchg'], 1, 1)}\% for ammonium, \cref{tab:p8chg}).  Seasonally, increases prior to upgrades were highest in the summer for nitrite/nitrate (\Sexpr{scinot(no_bf[3, 'perchg'], 1, 1)}\%) and in the fall for ammonium (\Sexpr{scinot(nh_bf[4, 'perchg'], 1, 1)}\%). Similarly, seasonal reductions post-upgrade were largest in the summer for nitrite/nitrate (\Sexpr{scinot(no_af[3, 'perchg'], 1, 1)}\%) and largest for ammonium in the winter (\Sexpr{scinot(nh_af[5, 'perchg'], 1, 1)}\%).

@@

% \section{References} % omit on submit
\begin{singlespace}
\bibliography{refs}
\end{singlespace}
\clearpage

%%%%%%
% figures

@@

<<clmchl, fig.height = 9.3, fig.width = 8, out.width = '0.9\\textwidth', cache = T, fig.cap = 'Trends in clam abundance and \\ac{chla} concentration from 1976 to 2013 at station D7 in Suisun Bay.  Invasion by \\textit{Potamocorbula amurensis} clams in the late 1980s and displacement of \\textit{Corbicula fluminea} was shown by changes in clam density (a, annual means), with biomass linked to salinity (b).  A decrease in \\ac{chla} concentration was also observed by changes in annual (c) and seasonal trends (f) based on \\ac{wrtds} results.  Reductions in \\ac{chla} concentration were coincident with an increase in SiO$_2$ and \\ac{din} concentrations (d, e), with the greatest increases in August (g, h). A significant ($p < 0.001$) relationship between clam biomass and \\ac{chla} concentration is shown in subfigure (i).  Flow relationships with \\ac{chla} concentration shown by \\ac{wrtds} have also changed over time (j, observations from June).'>>=
